Pyro -Velocity: Probabilistic RNA Velocity inference from single-cell data

### 2024-08-26 reproducibility/manuscript/v2.tex
chore(manuscript): update v2 diff file

We assume the dynamical gene expression is determined by the RNA
splicing process, and infer the unspliced and spliced gene expression
level from the differential equations proposed in velocyto
\citep{La_Manno2018-lj} and scVelo \citep{Bergen2020-pj} \begin{align}
\frac{d u\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \alpha^{\left(k_{cg}\right)}-\beta_g u\left(\tau^{\left(k_{cg}\right)}\right),
   \label{eq-dudt}\\
\frac{d s\left(\tau^{\left(k_{cg}\right)}\right)}{d \tau^{\left(k_{cg}\right)}}
  &= \beta_g u\left(\tau^{\left(k_{cg}\right)}\right)
  -\gamma_g s\left(\tau^{\left(k_{cg}\right)}\right). \label{eq-dsdt}
\end{align} In the equation, the subscript \(c\) is the cell dimension,
\(g\) is the gene dimension,
\(\left( u\left( \tau^{(k_{cg})} \right), s\left( \tau^{(k_{cg})} \right) \right)\)
are the unspliced and spliced expression functions given the change of
time per cell and gene. \(\tau_{cg}\) represents the displacement of
time per cell and gene with \begin{align}
 \tau^{(k_{cg})} &= \operatorname{softplus} \left( t_{c} - {t_{0}^{(k_{cg})}}_g \right) \\
 & = \log( 1 + \exp (t_c - {t_{0}^{(k_{cg})}}_g)), 
\end{align} in which \(t_c\) is the shared time per cell,
\({t_{0}^{(kcg)}}_g\) is the gene-specific switching time. Each cell and
gene combination has its transcriptional state
\(k_{cg} \in \{ 0, 1 \}\), where \(0\) indicates the activation state
and \(1\) indicates the expression state. Each gene has two switching
times for representing activation and repression: \({t_{0}^{(0)}}_g\) is
the first switching time corresponding to when the gene expression
starts to be activated, \({t_0^{(1)}}_g\) is the second switching time
corresponding to when the gene expression starts to be repressed. We
note that \(\alpha^{(1)}\) is shared for all the genes, while
\({\alpha^{(0)}}_g\) is learned independently for each gene.

The analytic solution of the differential equations to predict spliced
and unspliced gene expression given their parameters is derived by the
authors of scVelo and a theoretical RNA velocity study
\citep{Bergen2020-pj, Li2021-qa} and given in Eqs.
\ref{eq-solution-u}-\ref{eq-solution-s2}. \begin{align}
u\left(\tau^{\left(k_{c g}\right)}\right) 
  &= u_0^{\left(k_{c g}\right)}{ }_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} 
  \nonumber \\
&\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}
  {\beta_g}\left(1-e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right) 
  \label{eq-solution-u}\\
s\left(\tau^{\left(k_{c g}\right)}\right) 
  &= s_0^{\left(k_{c g}\right)} e^{-\gamma_g \tau^{\left(k_{c g}\right)}} 
  \nonumber \\
  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}{\gamma_g}
    \left(1-e^{-\gamma_g \tau^{\left(k_{c g}\right)}}\right) 
    \nonumber\\
  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}-\beta_g u_0^{\left(k_{c g}\right)}}
    {\gamma_g-\beta_g}\left(e^{-\gamma_g \tau^{\left(k_{c g}\right)}}
    -e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right), 
    \nonumber \\
  &\qquad \beta \neq \gamma \label{eq-solution-s} \\
s\left(\tau^{\left(k_{c g}\right)}\right) 
  &= {s_0^{\left(k_{c g}\right)}}_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} \nonumber \\
  &\hskip -24pt +\frac{\alpha^{\left(k_{c g}\right)}}{\beta_g}
  \left(1-e^{-\beta_g \tau^{(k c g)}}\right) 
  \nonumber \\
  &\hskip -24pt -\left(\alpha^{\left(k_{c g}\right)}
  -\beta_g u_0^{\left(k_{c g}\right)}{ }_g\right) \tau^{\left(k_{c g}\right)} 
  e^{-\beta_g \tau^{\left(k_{c g}\right)}}, \nonumber \\
  &\qquad \beta = \gamma \label{eq-solution-s2}
\end{align}

>>>>

We assume the dynamical gene expression is determined by the RNA
splicing process, and infer the unspliced and spliced gene expression
level from the ordinary differential equation (ODEs) proposed in
velocyto \citep{La_Manno2018-lj} and scVelo \citep{Bergen2020-pj}
\begin{align}
\frac{du}{dt} &= \alpha(t) - \beta u, \quad u(0) = u_0 \label{eq-dudt}\\
   \frac{ds}{dt} &= \beta u - \gamma s, \quad s(0) = s_0, \label{eq-dsdt}
\end{align} where \(u(t), s(t)\) are the unspliced and spliced
expression levels of a gene at time \(t\) under a transcription rate
\(\alpha(t)\) with possible temporal dependence, splicing rate
\(\beta\), and degradation rate \(\gamma\). We specify this model to a
setting that depends on cell \(c\) and gene \(g\) as follows:
\begin{align}
\frac{du_{cg}}{dt} &= \alpha_{cg}(t) - \beta_{g} u_{cg}, \quad u_{cg}(0) = u_{cg}^{(0)} \label{eq-dudt}\\
   \frac{ds_{cg}}{dt} &= \beta_{g} u_{cg} - \gamma_{g} s_{cg}, \quad s_{cg}(0) = s_{cg}^{(0)} \label{eq-dsdt}.
\end{align} In the equation, the subscript \(c\) is the cell dimension,
\(g\) is the gene dimension, \(\left( u_{cg}(t), s_{cg}(t) \right)\) are
the unspliced and spliced expression functions given the change of time
per cell and gene. We restrict attention to piecewise-constant
\(\alpha_{cg}(t)\) to capture gene-specific activation and repression.
We take special care to model a gene- and cell-specific switching time
that marks a single transition from activation to repression by
introducing a Bernoulli variable \(k_{cg}\) to model unknown activation
state. We assume our cell-by-gene data-matrix arrive as observations of
Poisson-counts related to the solution of the above ODEs at unknown
times \(\tau_{cg}\), which is modeled as a relationship between an
unknown latent time shared across each cell, \(t_c\), and unknown
gene-specific time-offsets \(t_{0,g}\) where all read counts for a
single cell occurred at an unknown, but shared latent time \(t_c\).
These relative times are also used to parametrize the Bernoulli process
for \(k_{cg}\). Importantly, we recognize that the initial conditions
are in fact unknown.

We propose and study two models: Model 1 assumes that spliced and
unspliced concentrations are both 0 at time 0; Model 2 considers these
initial conditions as unknowns with a log-Normal prior distribution. In
general, the solution space of ODEs becomes much richer when considered
over a domain of initial conditions (as opposed to a single point);
indeed, this affords Model 2 much greater expressivity. For clarity, we
first present the generative framework for both models, then provide
further interpretation and intuition.

First, we introduce the generative model that describes the various
unobserved times: \begin{align}
  % unit lognormal t_c
  t_c &\sim \text{LogNormal}(0, 1) \\
  % gene-specific t_0
  t^{(0)}_{0,g} &\sim \text{LogNormal}(0, 1) \\
  % switching time
  \Delta \textrm{switching}_g &\sim \text{LogNormal}(0, 1) \\
  % gene-specific t_1
  t^{(1)}_{0,g} &= t^{(0)}_{0,g} + \Delta \textrm{switching}_g \\
  %cell-gene-specific activation state
  k_{cg} &\sim \text{Bernoulli}(\textrm{logits}=t_c - t^{(1)}_{0,g}) \\
  % cell-gene-specific latent time
  \tau_{cg} &= \text{softplus}(t_c - t^{(k_{cg})}_{0,g}).
\end{align} Here, \(\tau_{cg}\) represents the displacement of time per
cell and gene with \begin{align}
 \text{softplus}(t) :=  \log( 1 + e^t).
\end{align} Recall that \(t_c\) is the shared time per cell,
\(t^{(k_{cg})}_{0,g}\) is the gene-specific switching time. Each cell
and gene combination has its transcriptional state
\(k_{cg} \in \{ 0, 1 \}\), where \(0\) indicates the activation state
and \(1\) indicates the expression state. Each gene has two switching
times for representing activation and repression: \(t^{(0)}_{0,g}\) is
the first switching time corresponding to when the gene expression
starts to be activated, \(t^{(1)}_{0,g}\) is the second switching time
corresponding to when the gene expression starts to be repressed, and is
determined by the first switching time and the gene-specific switching
time \(\Delta \text{switching}_g\). The cell-gene-specific activation
state \(k_{cg}\) is a Bernoulli random variable with logits equal to the
difference between the cell's shared time \(t_c\) and the time
\(t^{(1)}_{0,g}\) when the gene expression starts to be repressed.

Next we introduce the priors for the splicing parameters (where the
activation rate \(\alpha\) depends on the activation state \(k_{cg}\)
from above): \begin{align}
  \alpha^{(0)}_g &\sim \text{LogNormal}(0, 1) \\
  \beta_g &\sim \text{LogNormal}(0, 1) \\
  \gamma_g &\sim \text{LogNormal}(0, 1) \\
  \alpha_{cg} &= \begin{cases}
    \alpha^{(0)}_g & \text{if } k_{cg} = 0 \\
    0 & \text{if } k_{cg} = 1
  \end{cases}
\end{align}
\textbf{Note that $\alpha^{(1)}$ is shared for all the genes, while ${\alpha^{(0)}}_g$ is learned independently for
each gene. MATT: this was in the old text, but I think $\alpha^{(1)}$ is no longer used based on conversations with Alvin?}

Now, we describe the priors for the initial conditions, noting that this
is the only difference between Model 1 and Model 2: \begin{align}
  \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg} &\sim \begin{cases}
    (0, 0) & \text{Model 1} \\
    (\text{LogNormal}(0, 1), \text{LogNormal}(0, 1)) & \text{Model 2}
  \end{cases} \\
  u^{(0)}_{cg}, s^{(0)}_{cg} &= \begin{cases}
    \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg} & \text{if } k_{cg} = 0 \\
    \textrm{ODESolve}\Big( \hat{u}^{(0)}_{cg}, \hat{s}^{(0)}_{cg}, \alpha^{(0)}_g, \beta_g, \gamma_g; \ T_0=0, T_1=\Delta \textrm{switching}_g \Big) & \text{if } k_{cg} = 1
  \end{cases}
\end{align}

We define the ODE solution at time \(\tau_{cg}\) as: \begin{equation}
    \hat{u}_{cg}, \hat{s}_{cg} = \text{ODESolve}\Big( u^{(0)}_{cg}, s^{(0)}_{cg}, \alpha_{cg}, \beta_g, \gamma_g; \ T_0=0, T_1=\tau_{cg} \Big).
\end{equation}

Next, we define the observation model that gives rise to the observed
counts as: \begin{align}
  \mu^{(u)}_c &= \sum_{g=1}^G {u}^{\text{(obs)}}_{cg}, \quad \mu^{(s)}_c = \sum_{g=1}^G {s}^{\text{(obs)}}_{cg} \\
  \sigma^{(u)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( u_{cg}^{\text{(obs)}} - \mu^{(u)}_c \right)^2} \\
  \sigma^{(s)}_c &= \sqrt{\frac{1}{G} \sum_{g=1}^G \left( s_{cg}^{\text{(obs)}} - \mu^{(s)}_c \right)^2} \\
  \eta^{(u)}_c &\sim \text{Normal}\Big(\mu^{(u)}_c, \ \sigma^{(u)}_c\Big) \\
  \eta^{(s)}_c &\sim \text{Normal}\Big(\mu^{(s)}_c, \ \sigma^{(s)}_c\Big) \\
  \hat{\mu}^{(u)}_c &= \sum_{g=1}^G \hat{u}_{cg}, \quad \hat{\mu}^{(s)}_c = \sum_{g=1}^G \hat{s}_{cg} \\
  \lambda^{(u)}_{cg} &= \log(\hat{u}_{cg}) + \log(\eta^{(u)}_{c}) - \log(\hat{\mu}^{(u)}_c) \\
  \lambda^{(s)}_{cg} &= \log(\hat{s}_{cg}) + \log(\eta^{(s)}_{c}) - \log(\hat{\mu}^{(s)}_c) \\
  \hat{u}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(u)}_{cg})\Big) \\
  \hat{s}^{\text{(obs)}}_{cg} &\sim \text{Poisson}\Big(\exp (\lambda^{(s)}_{cg})\Big)
\end{align} Here, we use
\({u}^{\text{(obs)}}_{cg}, {s}^{\text{(obs)}}_{cg}\) to denote the
observed unspliced and spliced counts for cell \(c\) and gene \(g\). We
use \(\hat{u}^{\text{(obs)}}_{cg}, \hat{s}^{\text{(obs)}}_{cg}\) to
denote our generative model's prediction of these unspliced and spliced
expression levels. The generative process for modeling these observed
read counts given denoised gene transcript expression level
\(\hat{u}_{cg}, \hat{s}_{cg}\) considers the expected number of observed
reads for a given gene in a given cell as the number of transcripts
times the ratio of the cell's total reads to total transcripts.
\textbf{Improve descriptions of how noise is modeled in the observation model.}

\textbf{Need to update the analytic solutions, but first need to confirm the above is correct. Also, I recommend pushing all of the below analytic solutions to the appendix.}
The analytic solution of the differential equations to predict spliced
and unspliced gene expression given their parameters is derived by the
authors of scVelo and a theoretical RNA velocity study
\citep{Bergen2020-pj, Li2021-qa} and given in Eqs.
\ref{eq-solution-u}-\ref{eq-solution-s2}. \begin{align}
u\left(\tau^{\left(k_{c g}\right)}\right) 
  &= u_0^{\left(k_{c g}\right)}{ }_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} 
  \nonumber \\
&\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}
  {\beta_g}\left(1-e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right) 
  \label{eq-solution-u}\\
s\left(\tau^{\left(k_{c g}\right)}\right) 
  &= s_0^{\left(k_{c g}\right)} e^{-\gamma_g \tau^{\left(k_{c g}\right)}} 
  \nonumber \\
  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}}{\gamma_g}
    \left(1-e^{-\gamma_g \tau^{\left(k_{c g}\right)}}\right) 
    \nonumber\\
  &\hskip -24pt + \frac{\alpha^{\left(k_{c g}\right)}-\beta_g u_0^{\left(k_{c g}\right)}}
    {\gamma_g-\beta_g}\left(e^{-\gamma_g \tau^{\left(k_{c g}\right)}}
    -e^{-\beta_g \tau^{\left(k_{c g}\right)}}\right), 
    \nonumber \\
  &\qquad \beta \neq \gamma \label{eq-solution-s} \\
s\left(\tau^{\left(k_{c g}\right)}\right) 
  &= {s_0^{\left(k_{c g}\right)}}_g e^{-\beta_g \tau^{\left(k_{c g}\right)}} \nonumber \\
  &\hskip -24pt +\frac{\alpha^{\left(k_{c g}\right)}}{\beta_g}
  \left(1-e^{-\beta_g \tau^{(k c g)}}\right) 
  \nonumber \\
  &\hskip -24pt -\left(\alpha^{\left(k_{c g}\right)}
  -\beta_g u_0^{\left(k_{c g}\right)}{ }_g\right) \tau^{\left(k_{c g}\right)} 
  e^{-\beta_g \tau^{\left(k_{c g}\right)}}, \nonumber \\
  &\qquad \beta = \gamma \label{eq-solution-s2}
\end{align}
